A global sensitivity analysis of model uncertainty in aeroelastic wind turbine models

### 2020-01-22 Papers/Torque_Paper/article.tex
Ran simulation for paper

\begin{abstract}
Wind turbines are a complex multi-physics systems whose performance and life span is highly dependent on manufacturing, meteorological and operational factors. It is therefore important to quantify the impact of these parameters on the turbine response. We perform comprehensive sensitivity analysis of aeroelastic models typically employed to compute turbine response. This paper also seeks to quantify sensitivities due to manufacturing tolerance in the turbine blades. We propose a non-uniform rational basis splines approach to parametrize the geometry of the blade and further utilize this for quantifying geometric sensitivities at different radial locations. [Some important results]
\end{abstract}
\begin{keyword}
BEM, global sensitivity analysis, model uncertainty, Sobol indices
\end{keyword}
\end{frontmatter}

@@

\section{Introduction}
Aeroelastic models such as the Blade Element Momentum (BEM) models \cite{HandBook} play a critical role in the design, development and optimization of modern wind turbines. A large number of BEM models have been developed to predict turbine responses such as the structural loads and power output \cite{Vorpahl2013}.

As a consequence of the strong model assumptions at the basis of BEM theory, the results from BEM codes can be subject to significant inaccuracies or uncertainties. For example, the effect of sheared inflow \cite{Madsen2012} is not naturally accounted for in the theory, and needs to be incorporated via correction terms. Other major model uncertainties in BEM models are for example the time constant in dynamic stall models, the wake correction factor, the tip loss model parameter, and the lift- and drag-polars used to compute local aerodynamic forces. Especially for increasing turbine sizes, these model parameters  are foreseen to be not sufficiently accurate \cite{Sayed2019}. In other words, the uncertainty associated to the output of currently employed BEM models is rather large. 
% iIn order to arrive at robust BEM models that include an uncertainty estimate associated

@@

Clearly, the main novelty of the work lies in the study of the effect of model uncertainties rather than external conditions or geometric uncertainties. The current sensitivity analysis is part of the so-called WindTrue project, in which the long-term goal is the development of calibrated BEM models, that possess a quantified level of uncertainty. Anticipating on this goal, we are using the NM80 wind turbine model from the Tjaereborg wind farm, for which extensive measurement data is available through the DanAero project \cite{Troldborg2013} and IEA Task 29. The measurement data will enable the calibration step, and our sensitivity analysis will make clear which of the model parameters should be used in the calibration step.

@@

%A number of parameters describing meteorological and operational conditions as well as manufacturing specifications are needed as inputs for BEM simulations. In this work, we seek to quantify the sensitivities of these input parameters for different turbine responses. In the past a number of sensitivity studies have been performed to understand the influence of input parameters on different turbine response, for e.g. \cite{moriarty2002effect, eggers2003wind, McKay2014, dykes2014sensitivity, Robertson2018}. Many studies have confirmed that the wind parameters especially the wind speed and wind speed standard deviation have the most influence on aerodynamic performance of turbines. For example, in the case of an upstream turbine, the wind speed is most sensitive to power production. Furthermore, wind speed in combination with wind speed standard deviation has a large influence on the power production of turbines operating in the wake of an upstream turbine. Among operational factors, the rotor RPM is the most sensitive to power production. [More on parameter sensitivities on structural loads]
%
%[Sensitivity on the polars as novel element]
%
%[Mention this study is a part of IEA Task29 project]
%
%In this work, we also study the effect of manufacturing tolerances on the turbine response. Usually, there are some discrepancy between the  manufactured and nominally prescribed design of the turbine blade leading to a suboptimal performance. For BEM models, the turbine shape is described as a series of airfoils along the span of the blade where each airfoil shape is computed using the three quantities: chord length, thickness and twist. We perturb these three quantities to obtain a perturbed turbine blade and analyze with sections that have more influence on output quantities. 
%
%To compute parameter sensitivities we use the Sobol expansion approach, which decomposes the total variance of the quantity of interest into contributions from individual parameters and their combinations. To perform Sobol analysis, we use Matlab-based general purpose uncertainty quantification toolbox \texttt{UQLab} \cite{uqlab}. \texttt{UQLab}'s modular structure allows for easy integration with available BEM codes. For investigation the aeroelastic model of the  2MW NM80 turbine (with an 80m rotor) from the DANAERO project \cite{DANAERO} is utilized.

\section{Aeroelastic models}\label{sec:model_description}
\subsection{BEM model uncertainties}
State-of-the-art wind-turbine models simulate the aeroelastic behavior of wind turbines by combining the concept of momentum conservation of the flow (BEM) with the equations of motion for the structure, possibly extended with hydrodynamics of the sea and control algorithms \cite{Vorpahl2013}. In this work, we only concentrate on the first aspect, namely the prediction of flow and blade forces as given by the BEM method. The particular BEM code that we use is  ECN's \texttt{Aero-Module} \cite{Boorsma2012}.

For this case of a rigid turbine, with a uniform inflow field, the main model uncertainties that arise in the BEM formulation that we will discuss in this work are the following \cite{Hansen1993}:
\begin{itemize}
\item The use of (2D) \textit{airfoil polars}; uncertainty arises not only because the actual flow along the blade is 3D, but also because the 2D polars itself can be inaccurate, either when obtained from measurement or from 2D codes like \texttt{XFOIL};
%\item Sheared inflow
\item The assumption of locally 2D flow along the blade is also invalid at the tip; this is typically corrected via \textit{a tip loss model};
\item Three-dimensional flow effects also cause stall delay at the root; this is typically modelled via \textit{a dynamic stall model }and time constant;
\item Yawed inflow is not naturally included in BEM and typically incorporated via \textit{a wake correction}.

%\item Dynamic wake - Oye time lag constant
%\item Tower shadow constants – Potential flow / Downwind Empirical model

@@

\subsection{Geometric uncertainty}
We use Non-Uniform Rational Basis Splines (NURBS) \cite{rogers2000} to perturb the geometrical parameters of the turbine blade, such as the reference chord, twist or experimentally obtained drag-, lift-polar curves. The main advantage of using NURBS is that it provides great flexibility to approximate a large variety of curves with a limited number of control points which can be useful in reducing the number of uncertain parameter. %Further, the set of control points and knots can be directly manipulated to control the smoothness and curvature.
Use of NURBS has already been exploited in many aerodynamics applications, for instance, design optimization of wind-turbine blades, see for e.g. \cite{Ribeiro2012,Bottasso2014} and to parametrize chord and twist curves \cite{Echeverria2017}. In the following, we briefly outline the procedure used to parametrize curves using NURBS.
\subsubsection{NURBS based perturbation}
The value of NURBS curve at each location $x$ is computed using a weighted sum of $n$ basis functions (or B-splines):
\begin{equation}\label{NURB_curve}
S(x)  = \sum_{i=1}^{n} c_i B_{i,p}(x),
\end{equation}
where $S(x)$ is the value of the NURBS curve at location $x$; $c_i$ is the weight of the $i$-th control point and $B_{i,p}(x)$ is the value of B-spline corresponding to the $i$-th control point at $x$ with the subscript $p$ denoting its polynomial degree. A sequence of non-decreasing parameters $X = \{x_1,x_2,...,x_m\}$ called the \textit{knot vector} determines the domain in which any control point is active with the number of elements in the knot vector is given by $m = n+p+1$. B-splines are non-zero only within a knot interval $[x_i,x_{i+1})$ for $i = 1, 2, ..., m-1$ and are zero elsewhere. Therefore, only a few nearby control points are active when computing the NURBS curve \eqref{NURB_curve}. For a curve of polynomial degree $p$, there are $p+1$ active control points at any location $x$, for e.g., for a linear curve ($p=1$) only two adjacent control points are active at any location. Also, note that the number of B-splines are equal to the number of elements in $X$. Another useful property of the B-splines is that they are recursive in polynomial degree thus we can derive quadratic B-splines using linear B-splines, cubic from quadratic B-splines and so on. Starting with B-spline of degree 0 defined as 
\begin{equation}\label{linearBspline}
B_{i,0}(x) :=
\begin{cases}
1\quad x_i\leq x < x_{i+1},\quad i = 1,2,...,m,\\
0\quad\text{elsewhere,}
\end{cases} 
\end{equation}
we can derive higher order B-splines in the knot interval $[x_{i+p},x_{i+p+1})$ using the recurrence relation \cite{deBoor}:
\begin{equation}\label{NURBS_recurrence}
B_{i,p}(x) := \frac{x - x_i}{x_{i+p} - x_i}B_{i,p-1}(x) + \frac{x_{i+p+1}  -  x}{x_{i+p+1}  -  x_{i+1}}B_{i+1,p-1}(x),\quad p\geq1.
\end{equation} 
These B-splines can be constructed efficiently using the de Boor's algorithm \cite{deBoor}. Note that to perform above computation, we need to pad the knot vector $X$ by repeating the first and last elements $p$ times each. In Fig. \ref{B-splines}, we show linear, quadratic and cubic splines for $x\in[0,1]$ [Also show the padding].

Next, we describe steps to generate perturbed samples of chord from a given reference chord, $S_{ref}(x)$, using NURBS based parametrization. The first step is to approximate the given reference chord using a NURBS curve with a fixed degree $p$ and a limited number of knots $X =\{x_j\}_{j=0}^n$. The knots can be uniformly spaced or chosen heuristically, more advanced approaches for knot selection can be found in, for e.g. \cite{razdan1999knot,LI2005791}. Next we sample $S_{ref}(x)$ at $X$ and also compute $B_{i,p}(x_j), \text{ for }i, j = 1,2, ..., n$.  Now we can formulate a linear system to solve for a set of control points $\mathbf{c}=\{c_i\}_{i=0}^n$ as:
\begin{equation}\label{nurbs_inversion}
\mathbf{B}\mathbf{c} = \mathbf{S}_{ref},
\end{equation}
where $\mathbf{S}_{ref}\in\mathbb{R}^{n}$ is a vector containing sampled values of $S_{ref}(x)$ and $\mathbf{B}\in \mathbb{R}^{n\times n}$ is a matrix with $j$-th row consisting of values of $n$ B-splines sampled at the location $x_j$. Once the control points are obtained, we can derive the approximate reference curve $S_n(x)$ using \eqref{NURB_curve}. In this work, we use uniform random variable to perturb the baseline control points $\mathbf{c}$.

There are two ways to obtain perturbed curves either by independently perturbing each of the control points (local perturbation) or using the same random number to perturb all the control points (global perturbation). We use local perturbation for chord control points Ch1-Ch5 and twist control points Tw1-Tw6 control points, see Fig. \ref{perturbed_samples} (a) - (b), respectively. Each control points are perturbed using a uniform random number with bounds given by $\pm5\%$ of the magnitude of control point. For lift and drag curve, all control points are perturbed using same uniform random number resulting in samples of the lift and drag curves as shown in Fig. \ref{perturbed_samples} (c) - (d), respectively. For simplicity, we only consider uniform distributions for perturbing the control points, however, we can also use other probability distributions for this. Also note that when using a uniform distribution with bounds given by relative magnitude of control points, we tend to obtain small perturbations for control points with values close to zero.
\begin{figure}[h!]
\centering
\includegraphics[trim={0 1.6cm 0 1.5cm},clip, scale=0.67]{figure1.pdf}
\caption{Random realization of chord, twist, lift- (Cl2) and drag-polars (Cd2).}
\label{perturbed_samples}
\end{figure}

\section{Global sensitivity analysis}\label{sec:GSA}
The objective of sensitivity analysis is to quantify the relative significance of individual inputs (or in combination) and how variations in input values affects the output of interest. In engineering, sensitivity analysis can be employed for a number of reasons: to determine the stability and robustness of a computational model with respect to input parameters, for simplification of stochastic models by fixing the insensitive parameters, and to guide data acquisition campaigns and experimental design to refine the data on sensitive parameters. Sensitivity analysis techniques can be classified as local and global methods, see \cite{RSmith}. In local sensitivity analysis, individual parameters are perturbed around their nominal values allowing for the description of output variability only in a small neighbourhood of nominal input values. Although, local approaches are widely employed due to their ease of implementation and low computational cost, they are unable to quantify global behavior of nonlinearly parametrized models such as aeroelastic models. Global sensitivity approaches, on the other hand, consider the entire range of input values to compute output sensitivities. Therefore, global sensitivity analysis is more suitable for aeroelastic models considered in this work.

@@

The PCE $f^{K}(\mathbf{z})$ of the computational model $Y$ is defined as a weighted sum of multivariate polynomials in $\mathbf{z}$ \cite{RSmith}
Y = f^{K}(\mathbf{z}) = \sum_{|\mathbf{k}| = 0}^K y_{\mathbf{k}}\Psi_{\mathbf{k}}(\mathbf{z}),

@@

where $\mathbf{k}\in  \mathbb{N}_0^M$ is a $M$-dimensional multi-index with the magnitude $|\mathbf{k}| = k_1+k_2 + ... + k_M$, whereas $\Psi_{\mathbf{k}}(\mathbf{z})$ is the multivariate orthogonal polynomials that is computed using a tensor product of univariate polynomials $\psi_{k_i}^{(i)}(z_i)$

@@

The choice of the univariate orthogonal polynomial $\psi_{k_i}$ depends on the type of random variable, $z_i$. For example, if for uniformly distributed random variable, the Legendre family of polynomials is used, see  \cite{Xiu2002} for details. Projection methods are typically employed for the calculation

Quadrature based integration methods suffer from curse of dimensionality where the number of integration points rises exponentially with an increase in the number of dimensions. For applications with a fixed computational budget methods that minimizes the approximation error with a given number of samples are usually more practical. Next, we briefly describe two such methods to compute PCE coefficients based on a given number of model evaluations.

\subsubsection{Ordinary Least Square}
The 
\subsubsection{Least Angle Regression (PCE\_LAR)}
\subsection{Sensitivity analysis with discrete random variable}
%\section{Description of test case}
%\section{Sensitivity analysis workflow}
\section{Results}\label{sec:results}

\subsection{NM80}
 and the turbine is the 2MW NM80 turbine from the DANAERO project \cite{Troldborg2013} with a blade radius of 38.8m. The data for lift ($Cl$) and drag ($Cd$) polars are available at four locations along the blade radius at 11.87m, 17.82m, 28.97m and 35.53m. The reference value of polars are obtained from wind-tunnel experiment with 3D corrections. Random samples of chord, twist, lift- and drag-polars are obtained by perturbing the control points with a uniformly distributed random variable. For chord and twist, we independently perturb control points at different locations as shown in Fig. \ref{fig:samples} (a) - (b). For $Cl$ and $Cd$, all control points are perturbed using a same random number.

\subsubsection{Verification of GSA method}
Convergence study of UQLab with increasing number of samples for Quad, OLS, LARS.

>>>>

\begin{abstract}
Sensitivity analysis of model parameters associated with the Blade element momentum models is studied. Sobol based 
\end{abstract}
\begin{keyword}
BEM, global sensitivity analysis, model uncertainty, Sobol indices
\end{keyword}
\end{frontmatter}

@@

\section{Introduction}
Aeroelastic models such as the Blade Element Momentum (BEM) models \cite{HandBook} play a critical role in the design, development, and optimization of modern wind turbines. A large number of BEM models have been developed to predict turbine responses such as the structural loads and power output \cite{Vorpahl2013}.

As a consequence of the strong model assumptions at the basis of BEM theory, the results from BEM codes can be subject to significant inaccuracies or uncertainties. For example, the effect of sheared inflow \cite{Madsen2012} is not naturally accounted for in the theory and needs to be incorporated via correction terms. Other major model uncertainties in BEM models are for example the time constant in dynamic stall models, the wake correction factor, the tip loss model parameter, and the lift- and drag-polars used to compute local aerodynamic forces. Especially for increasing turbine sizes, these model parameters are foreseen to be not sufficiently accurate \cite{Sayed2019}. In other words, the uncertainty associated with the output of currently employed BEM models is rather large. 
% to arrive at robust BEM models that include an uncertainty estimate associated

@@

The main novelty of the work lies in the study of the effect of model uncertainties rather than external conditions or geometric uncertainties. The current sensitivity analysis is part of the so-called WindTrue project, in which the long-term goal is the development of calibrated BEM models, that possess a quantified level of uncertainty. Anticipating on this goal, we are using the NM80 wind turbine model from the Tjaereborg wind farm, for which extensive measurement data is available through the DanAero project \cite{Troldborg2013} and IEA Task 29. The measurement data will enable the calibration step, and our sensitivity analysis will make clear which of the model parameters should be used in the calibration step.

@@

% Several parameters describing meteorological and operational conditions, as well as manufacturing specifications, are needed as inputs for BEM simulations. In this work, we seek to quantify the sensitivities of these input parameters for different turbine responses. In the past, several sensitivity studies have been performed to understand the influence of input parameters on different turbine response, e.g. \cite{moriarty2002effect, eggers2003wind, McKay2014, dykes2014sensitivity, Robertson2018}. Many studies have confirmed that wind parameters especially wind speed and wind speed standard deviation have the most influence on the aerodynamic performance of turbines. For example, in the case of an upstream turbine, the wind speed is most sensitive to power production. Furthermore, wind speed in combination with wind speed standard deviation has a large influence on the power production of turbines operating in the wake of an upstream turbine. Among operational factors, the rotor RPM is the most sensitive to power production. [More on parameter sensitivities on structural loads]
%
%[Sensitivity on the polars as novel element]
%
%[Mention this study is a part of IEA Task29 project]
%
%In this work, we also study the effect of manufacturing tolerances on the turbine response. Usually, there is some discrepancy between the manufactured and nominally prescribed design of the turbine blade leading to suboptimal performance. For BEM models, the turbine shape is described as a series of airfoils along the span of the blade where each airfoil shape is computed using the three quantities: chord length, thickness, and twist. We perturb these three quantities to obtain a perturbed turbine blade and analyze with sections that have more influence on output quantities. 
%
%To compute parameter sensitivities we use the Sobol expansion approach, which decomposes the total variance of the quantity of interest into contributions from individual parameters and their combinations. To perform Sobol analysis, we use Matlab-based general-purpose uncertainty quantification toolbox \texttt{UQLab} \cite{uqlab}. \texttt{UQLab}'s modular structure allows for easy integration with available BEM codes. For investigation, the aeroelastic model of the 2MW NM80 turbine (with an 80m rotor) from the DANAERO project \cite{DANAERO} is utilized.

\section{Aeroelastic models}\label{sec:model_description}
\subsection{BEM model uncertainties}
State-of-the-art wind-turbine models simulate the aeroelastic behavior of wind turbines by combining the concept of momentum conservation of the flow (BEM) with the equations of motion for the structure, possibly extended with the hydrodynamics of the sea and control algorithms \cite{Vorpahl2013}. In this work, we only concentrate on the first aspect, namely the prediction of flow and blade forces as given by the BEM method. The particular BEM code that we use is ECN's \texttt{Aero-Module} \cite{Boorsma2012}.

For this case of a rigid turbine, with a uniform inflow field, the main model uncertainties that arise in the BEM formulation that we will discuss in this work are the following \cite{Hansen1993}:
\begin{itemize}
\item The use of (2D) \textit{airfoil polars}; uncertainty arises not only because the actual flow along the blade is 3D, but also because the 2D polars itself can be inaccurate, either when obtained from measurement or 2D codes like \texttt{XFOIL};
%\item Sheared inflow
\item The assumption of locally 2D flow along the blade is also invalid at the tip; this is typically corrected via \textit{a tip loss model};
\item Three-dimensional flow effects also cause stall delay at the root; this is typically modeled via \textit{a dynamic stall model }and time constant;
\item Yawed inflow is not naturally included in BEM and typically incorporated via \textit{a wake correction}.

%\item Dynamic wake - Oye time lag constant
%\item Tower shadow constants " Potential flow / Downwind Empirical model

@@

\subsection{Geometric uncertainty}
We use Non-Uniform Rational Basis Splines (NURBS) \cite{rogers2000} to perturb the geometrical parameters of the turbine blade, such as the reference chord, twist or experimentally obtained drag-, lift-polar curves. The main advantage of using NURBS is that it provides great flexibility to approximate a large variety of curves with a limited number of control points which can be useful in reducing the number of uncertain parameters. %Further, the set of control points and knots can be directly manipulated to control the smoothness and curvature.
Use of NURBS has already been exploited in many aerodynamics applications, for instance, design optimization of wind-turbine blades, see for e.g. \cite{Bottasso2014,Ribeiro2012} and to parametrize chord and twist curves \cite{Echeverria2017}. In the following, we briefly outline the procedure used to parametrize curves using NURBS.
\subsubsection{NURBS based perturbation}
The value of NURBS curve $S(x)$ is computed using a weighted sum of $n$ basis functions (or B-splines):
\begin{equation}\label{NURB_curve}
S(x) = \sum_{i=1}^{n} c_i B_{i,p}(x),
\end{equation}
where $c_i$ is the weight of the $i$-th control point and $B_{i,p}(x)$ is the value of B-spline corresponding to the $i$-th control point at $x$ with the subscript $p$ denoting its polynomial degree. A sequence of non-decreasing parameters $X = \{x_1,x_2,...,x_m\}$ called the \textit{knot vector} determines the domain in which any control point is active with the number of elements in the knot vector is given by $m = n+p+1$. A B-spline $B_{i,p}(x)$ is non-zero only within a knot interval $[x_i,x_{i+1})$ for $i = 1, 2, ..., m-1$ and are zero elsewhere. Therefore, only a few nearby control points are active when computing the NURBS curve \eqref{NURB_curve}. For a curve of polynomial degree $p$, there are $p+1$ active control points at any location $x$, for e.g., for a linear curve ($p=1$) only two adjacent control points are active at any location. Also, note that the number of B-splines are equal to the number of elements in $X$. Another useful property of the B-splines is that they are recursive in polynomial degree thus we can derive quadratic B-splines using linear B-splines, cubic from quadratic B-splines and so on. Starting with B-spline of degree 0 defined as 
\begin{equation}\label{linearBspline}
B_{i,0}(x) :=
\begin{cases}
1\quad x_i\leq x < x_{i+1},\quad i = 1,2,...,m,\\
0\quad\text{elsewhere,}
\end{cases} 
\end{equation}
we can derive higher order B-splines in the knot interval $[x_{i+p},x_{i+p+1})$ using the recurrence relation \cite{deBoor}:
\begin{equation}\label{NURBS_recurrence}
B_{i,p}(x) := \frac{x - x_i}{x_{i+p} - x_i}B_{i,p-1}(x) + \frac{x_{i+p+1} - x}{x_{i+p+1} - x_{i+1}}B_{i+1,p-1}(x),\quad p\geq1.
\end{equation} 
These B-splines can be constructed efficiently using the de Boor's algorithm \cite{deBoor}. Note that to perform above computation, we need to pad the knot vector $X$ by repeating the first and last elements $p$ times each. In Fig. \ref{B-splines}, we show linear, quadratic and cubic splines for $x\in[0,1]$ [Also show the padding].

Next, we describe steps to generate perturbed samples of chord from a given reference chord, $S_{ref}(x)$, using NURBS based parametrization. The first step is to approximate the given reference chord using a NURBS curve with a fixed degree $p$ and a limited number of knots $X =\{x_j\}_{j=0}^n$. The knots can be uniformly spaced or chosen heuristically, more advanced approaches for knot selection can be found in, for e.g. \cite{LI200579,razdan1999knot}. Next we sample $S_{ref}(x)$ at $X$ and also compute $B_{i,p}(x_j), \text{ for }i, j = 1,2, ..., n$. Now we can formulate a linear system to solve for a set of control points $\mathbf{c}=\{c_i\}_{i=0}^n$ as:
\begin{equation}\label{nurbs_inversion}
\mathbf{B}\mathbf{c} = \mathbf{S}_{ref},
\end{equation}
where $\mathbf{S}_{ref}\in\mathbb{R}^{n}$ is a vector containing sampled values of $S_{ref}(x)$ and $\mathbf{B}\in \mathbb{R}^{n\times n}$ is a matrix with $j$-th row consisting of values of $n$ B-splines sampled at the location $x_j$. Once the control points are obtained, we can derive the approximate reference curve $S_n(x)$ using \eqref{NURB_curve}. In this work, we use uniform random variable to perturb the baseline control points $\mathbf{c}$.

There are two ways to obtain perturbed curves either by independently perturbing each of the control points (local perturbation) or using the same random number to perturb all the control points (global perturbation). We use local perturbation for chord control points Ch1-Ch5 and twist control points Tw1-Tw6 control points, see Fig. \ref{perturbed_samples} (a) - (b), respectively. Each control point are perturbed using a uniform random number with bounds given by $\pm5\%$ of the magnitude of a control point. For lift and drag curves, all control points are perturbed using the same uniform random number resulting in samples of the lift and drag curves as shown in Fig. \ref{perturbed_samples} (c) - (d), respectively. For simplicity, we only consider uniform distributions for perturbing the control points, however, we can also use other probability distributions for this. Also note that when using a uniform distribution with bounds given by the relative magnitude of control points, we tend to obtain small perturbations for control points with values close to zero.
\begin{figure}[h!]
\centering
\includegraphics[trim={0 0cm 0 0cm},clip, scale=.75]{basis_example.eps}
\caption{}
\label{perturbed_samples}
\end{figure}

\section{Global sensitivity analysis}\label{sec:GSA}
The objective of sensitivity analysis is to quantify the relative significance of individual inputs (or in combination) and how variations in input values affect the output of interest. In engineering, sensitivity analysis can be employed for several reasons: to determine the stability and robustness of a computational model for input parameters, for simplification of stochastic models by fixing the insensitive parameters, and to guide data acquisition campaigns and experimental design to refine the data on sensitive parameters. Sensitivity analysis techniques can be classified as local and global methods, see \cite{RSmith}. In a local sensitivity analysis, individual parameters are perturbed around their nominal values allowing for the description of output variability only in a small neighborhood of nominal input values. Although, local approaches are widely employed due to their ease of implementation and low computational cost, they are unable to quantify the global behavior of nonlinearly parametrized models such as aeroelastic models. Global sensitivity approaches, on the other hand, consider the entire range of input values to compute output sensitivities. Therefore, global sensitivity analysis is more suitable for aeroelastic models considered in this work.

@@

The PCE approximation $f^{K}(\mathbf{z})$ of the computational model $Y$ can be defined as a weighted sum of multivariate polynomials in $\mathbf{z}$ \cite{RSmith}
\begin{equation}\label{eq:PCE}
f(\mathbf{z}) \approx f^{K}(\mathbf{z}) = \sum_{|\mathbf{k}| = 0}^K w_{\mathbf{k}}\Psi_{\mathbf{k}}(\mathbf{z}),

@@

where $\mathbf{k}\in \mathbb{N}_0^M$ is a $M$-dimensional multi-index with the magnitude $|\mathbf{k}| = k_1+k_2 + ... + k_M$. The total number of terms $K$ in \eqref{eq:PCE} can be based on a standard truncation scheme where we use all polynomials in $M$ random variable with order less than or equal to $p$. More advanced truncation schemes such as the maximum interaction or a hyperbolic scheme can be employed depending on the application \cite{BlatmanThesis}. The multivariate polynomials $\Psi_{\mathbf{k}}(\mathbf{z})$ is orthogonal with respect to the joint probability distribution of random vector $\mathbf{z}$. The multivariate distribution is computed using a tensor product of univariate polynomials $\psi_{k_i}^{(i)}(z_i)$

@@

The choice of the univariate orthogonal polynomial $\psi_{k_i}$ depends on the type of random variable, $z_i$. For example, for uniform random variable the Legendre family of polynomials is used, see \cite{Xiu2002} for details. Exploiting the orthogonality of $\Psi_{\mathbf{k}}(\mathbf{z})$, the coefficients can be computed using the projection approach:
\begin{equation}\label{eq:projection}
w_{\mathbf{k}} = \mathbb{E}[\Psi_{\mathbf{k}}(\mathbf{z})\cdot f^{K}(\mathbf{z})]
The above expectation can be computed using a quadrature based integration methods such as the Gaussian quadrature, sparse quadrature rules (for e.g., Smolyak's tensor grid), etc. Quadrature based integration methods suffer from curse of dimensionality where the number of integration points rises exponentially with an increase in the number of dimensions. In the current work where we have a fixed computational budget, methods that minimizes the approximation error for a given number of samples are more practical. Next, we briefly describe two such methods to compute PCE coefficients based on a fixed number of model evaluations.
\subsubsection{Ordinary Least Square}
Instead of using projection based method one can also use least squares based approaches to compute coefficients \cite{Berveiller2006}. The PCE expansion \eqref{eq:PCE} can be expressed as a vector-matrix product
\begin{equation}\label{eq:LSM}
Y = f^{K}(\mathbf{z}) +\varepsilon_K = \sum_{|\mathbf{k}| = 0}^K w_{\mathbf{k}}\Psi_{\mathbf{k}}(\mathbf{z}) + \varepsilon_K = \mathbf{w}^T\boldsymbol{\Psi}(\mathbf{z}) +\varepsilon_K,
where $\mathbf{w} = \{w_0,w_1,...,w_K\}^T$ is the coefficient vector, $\boldsymbol{\Psi}(\mathbf{z}) = \{\Psi_0(\mathbf{z}), \Psi_1(\mathbf{z}), ..., \Psi_K(\mathbf{z})\}^T$ is the matrix containing multivariate polynomials in $\mathbf{z}$ and $\varepsilon_K$ is the approximation error. The least-square minimization problem can then be formulated as:
\begin{equation}\label{eq:PCE_OLS}
\mathbf{\hat{w}} = \text{arg min }\mathbb{E}\left[\left(\mathbf{w}^T \boldsymbol{\Psi}(\mathbf{z}) - f(\mathbf{z})\right)^2\right]
The coefficient vector $\mathbf{\hat{w}}$ can be computed using the Ordinary Least Square (OLS) method. Given $N$ model evaluations $\mathbf{Y} = \{f(\mathbf{z}_1), f(\mathbf{z}_2), ..., f(\mathbf{z}_N)\}$, we can obtain the coefficients as:
\begin{equation}\label{eq:OLS}
\hat{\mathbf{w}} = (\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbf{Y},
where $\mathbf{X}\in \mathbb{R}^{N\times K}$ is the design matrix where $\mathbf{X}_{i,j} = \Psi_{j}(\mathbf{z}_i)$, $i = 1,2, ..., N$ and $j = 0,1,2,..., K$. Although the OLS method allows to compute the coefficients using a limited number of model evaluations one the main drawback of this approach is that it may result in a PCE model consisting of high-order interaction terms that may lead to overfitting. For many engineering problems, low-order interactions between random input variables are more important. Therefore, the least square form in \eqref{eq:PCE_OLS} can be regularized such that a the minimization results in a low order sparse PCE model.

\subsubsection{Least Angle Regression}
Least angle regression (LARS) is one of the several methods can be applied to compute sparse PCE models \cite{BlatmanThesis,BLATMAN2011}.

The LARS algorithm can be summarized in the following steps:

\subsection{Sensitivity analysis with discrete random variable}

\section{Results}\label{sec:results}

\subsection{NM80}
 and the turbine is the 2MW NM80 turbine from the DANAERO project \cite{Troldborg2013} with a blade radius of 38.8 m. The data for lift ($Cl$) and drag ($Cd$) polars are available at four locations along the blade radius at 11.87 m, 17.82 m, 28.97 m, and 35.53 m. The reference value of polars is obtained from the wind-tunnel experiment with 3D corrections. Random samples of chord, twist, lift- and drag-polars are obtained by perturbing the control points with a uniformly distributed random variable. For chord and twist, we independently perturb control points at different locations as shown in Fig. \ref{fig:samples} (a) - (b). For $Cl$ and $Cd$, all control points are perturbed using the same random number.

\subsubsection{Verification of GSA method}
Convergence study of UQLab with an increasing number of samples for OLS and LARS.
\begin{figure}[h!]
\centering
\includegraphics[trim={0cm 0cm 0 0cm},clip, scale=1.3]{nurbs_approx.eps}
\caption{Random realization of chord, twist, lift- (Cl2) and drag-polars (Cd2).}
\label{perturbed_samples}
\end{figure}

\begin{figure*}[h!]
    \centering
    \begin{subfigure}[t]{0.32\textwidth}
        \centering
        \includegraphics[trim={1cm 0cm 0 0cm},clip, scale=0.22]{convergence_chord1_4.eps}{a}
        \caption{$M=4$}
    \end{subfigure}%
    ~ 
    \begin{subfigure}[t]{0.32\textwidth}
        \centering
        \includegraphics[trim={1cm 0cm 0 0cm},clip, scale=0.22]{convergence_chord1_6.eps}
        \caption{$M=6$}
    \end{subfigure}
      ~ 
    \begin{subfigure}[t]{0.32\textwidth}
        \centering
        \includegraphics[trim={1cm 0cm 0 0cm},clip, scale=0.22]{convergence_chord1_8.eps}
        \caption{ $M=8$}
 \end{subfigure}
    \caption{ Convergence of $PCE\_OLS$ and $PCE\_LARS$ with increasing number of uncertain inputs $M$ and number of samples $N$}
\end{figure*}

\begin{figure}[h!]
\centering
\includegraphics[trim={1cm 0cm 0 0cm},clip, scale=0.67]{figure1.pdf}
\caption{Random realization of chord, twist, lift- (Cl2) and drag-polars (Cd2).}
\label{perturbed_samples}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[trim={1cm 0cm 0 0cm},clip, scale=0.4]{SA_Power_chord_twist_Cl_Cd.eps}
\caption{}
\label{perturbed_samples}
\end{figure}
